---
title: "Del 2: What effect does baseline myonuclear number have on type-I and type-II myofiber hypertrophy following a 6-week progressive volume RT-protocol?"
author: "Ketil Barstad"
date: 2025-11-18
bibliography: references.bib
csl: vancouver.csl
link-citations: true
format:
  pdf:
    pdf-engine: xelatex
    mainfont: "TeX Gyre Pagella"
toc: true
toc-title: Innholdsfortegnelse
toc-depth: 3
toc-expand: true
number-sections: true
---

```{r}
#Last inn pakkene og datasettet vi skal bruke
library(tidyverse)
library(exscidata)
library(car)
view(hypertrophy)
```

```{r}

hypertrophy |>
  select(PARTICIPANT, contains("NUCLEI"), contains("CSA"), contains("VL_")) |>
  summary()

m1 <- lm(VL_T3 ~ VL_T1 + FAST_NUCLEI_T1 + SLOW_NUCLEI_T1, data = hypertrophy)

summary(m1)

vif(m1)

```

# Introduction

Skeletal muscles contain hundreds to thousands of myonuclei, each of them governing a finite amount of cytpolasms within the myofiber

Our research question: Does baseline myonuclear density predict the hypertrophic response in the quadriceps following a 6-week progressive volume RT-protocol?

Alternative hypothesis: Baseline myonuclear density predicts the hypertrophic response following a 6-week progressive volume RT-protocol. Due to the heterogeneity of the studies in the field, we are not assuming a directionality of the effect. This study is exploratory, and intended for hypothesis generation. 

# Methods 

## Baseline characteristics of participants

## Resistance-training protocol 

## Myonuclear number determination

# Statistical analysis

## Rationale for use of model 

# Results

# Discussion

## Limitations 

# Konklusjon